---
title: \normalsize Utilization of Antiemetic Medication as a Marker of Healthcare
  Disparities in Anesthesia \newline A Bayesian Hierarchical Model Using the National
  Anesthesia Clinical Outcomes Registry
csl: Bibliography/anesthesia-and-analgesia.csl
fontfamily: pslatex
output:
  pdf_document:
    includes:
      in_header: header.tex
    keep_tex: yes
  word_document: default
geometry: margin=1in
header-includes: \setlength\parindent{24pt}
linestretch: 2
fontsize: 12pt
bibliography: Bibliography/AQI_bibliography6May2015.bib
---

\setlength{\parindent}{0ex}

# Authors

Michael H. Andreae, MD, 

Department of Anesthesiology, Montefiore Medical Center, Albert Einstein College of Medicine

Charles Hall, PhD, 

Department of Department of Epidemiology & Population Health, Albert Einstein College of Medicine

# Corresponding author:

Michael Andreae, MD, Email: mandreae@montefiore.org, Tel: +1 (718)920 2802, Address: Department of Anesthesiology, Montefiore Medical Center, 111 E 210th Street, Bronx, NY 10467, USA

# Acknowledgement

This research is supported in part by the National Center for Advancing Translational Sciences (NCATS), a component of the National Institutes of Health (NIH), through CTSA grant numbers 5KL2TR001071-03 and UL1 TR001073. Its contents are solely the responsibility of the authors and do not necessarily represent the official views of the NIH.

\pagebreak

\textbf{Abstract}

US Health care disparities persist despite repeated countermeasures. Research identified race, ethnicity, gender, and socio-economic status as predictors, mediated through individual provider and/or systemic biases; little research exists in anesthesiology. We chose antiemetic prophylaxis as a surrogate marker for anesthesia quality by individual providers, because antiemetics are universally available, indicated independent of patient co-morbidities and not yet impacted by regulatory or financial constraints. We hypothesize that socioeconomic markers (like insurance status or median income in the patients home zip code area) predict antiemetic prophylaxis as a marker of anesthesia quality. We fit a mixed logistic regression models in the subset of anesthesia cases in the National Anesthesia Clinical Outcomes Registry (NACOR) with complete electronic anesthesia records to test our hypothesis. Seven institutions reported insurance status and antiemetic utilization for 176,334 anesthesia cases, out of the 12 million cases in NACOR 2014. Bivariate analysis, multivariate logistic regression and our Bayesian hierarchical model all showed a strong and statistically significant association between socioeconomic markers and the odds of receiving antiemetic prophylaxis; **for medicaid versus commercially insured patients, the odds ratio of receiving the antiemetic ondansetron is 0.63 in our final logistic regression model with a 95% confidence interval of [0.62, 0.66] with similar Bayesian inferences.** Our data support the notion that individual anesthesia providers administer lesser quality care to patients with lower socioeconomic status. We controlled for some important patient characteristics as well as for procedure and provider. Our sample may not be representative and may fail to account for all confounders. Our results point to possible unappreciated healthcare disparities in anesthesia at the provider level which are worrisome and deserve further investigation and vigorous countermeasures.

```{r, global_options, echo=FALSE}
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

\newpage

\setcounter{secnumdepth}{2}
\setlength{\parindent}{10ex}

# Introduction

## Healthcare disparities in the United States
The healthcare disparities in the United States of America described decades ago by Gornick[@Gornick_8703185], unabatedly persist to the present[@Schoen_24226092] and are linked to social determinants of health and equality[@Cooper_22566148]. Health inequality is worse in the US than abroad [@Schoen_24226092]. Strikingly, life expectancy for African Americans is almost 4 years shorter compared to the Caucasian American population[@Kochanek_24152376]. The medical profession cannot and will not accept this grave inequality which contradicts basic principles of medical ethics[@IOM_2003]. Poverty, poor education, differences in medical insurance coverage, geographic location, legal or social status, race & gender, patient and community attitudes & perceptions and, last but not least, provider bias might lead to healthcare disparities also in anesthesia[@Silber_23719571]. This paper investigates if antiemetic utilization as a marker of high quality anesthesia care is predicted by medical insurance status in the large National Anesthesia Clinical Outcomes Registry (NACOR)[@Dutton_21871404]. 

### Known predictors of healthcare disparities

#### Race and ethnicity impact survival. 
African-Americans have been shown in innumerable studies[@IOM_2003] to have inferior health outcomes compared to their White contemporaries[@Nelson_12152921]. Race is associated with lower socioeconomic status, income and education . However, the observed associations persist after controlling for these material factors[@Santry_22269267]. Hispanic minorities likewise experience a systematic gap in health outcomes[@Juckett_23317025]. 
Gender and sexual orientation determine treatment and access. Gender and sexual orientation can also lead to systematic bias and consequently to inferior treatment and outcomes[@Schoen_24226092]. Homo- and bisexual men showed poorer health than heterosexual controls; lesbian and bisexual women likewise were at increased risk for heart disease[@Fredriksen-Goldsen_23763391]. Leresche discusses the exemplary observation that women experience higher pain scores, cautioning that this may be confounded by lower pain thresholds[@Leresche_21210309]. As another example, gender differences exist in orthopedics with regards to osteoporosis, fracture risk and utilization of services[@Cawthon_21264553]. Women experience in general more difficult access to care, the more so the poorer and less educated they are[@Schoen_12627610]. 

### Socioeconomic status determines health. 
Low socioeconomic status is associated with higher mortality in many populations. Education, income, legal status, geographic location, social networks may have different causal pathways leading to inferior health outcomes.

#### Income and education predict mortality and outcomes. 
Sabanayagam tried to compare income and education as predictors of mortality and found the poverty income ratio a better predictor of mortality in a national multi-ethnic sample of US adults[@Sabanayagam_22524123]. Education affects health outcomes. Berkman recently reviewed the literature and concluded that low health literacy is associated with inferior utilization of health care services and worse clinical outcomes[@Berkman_21768583].

#### Treatment and outcomes vary according to geographic location. 
Geographic differences in health outcomes occur at the state level, but can go down to the county and zip code: Differences in healthcare delivery systems across states can lead to variance in health outcomes beyond the influence of poverty[@McCarthy_24056035]. Lower average income in constituents sharing the same zip code was associated with higher disability and greater hospital utilization[@Cooper_22566148]. Rates of certain surgical procedures varied within the same Los Angeles County; much of this variability may be due to different ethnic mixes in the varied neighborhoods rather than differences in health status alone. This geographic variability is important when investigating large datasets drawing data from across the nation, like the NACOR dataset we propose to study.

#### Immigration and legal status determine health status. 
Page-Reeves demonstrated this eloquently for diabetes care in immigrants living in New Mexico[@Page-Reeves_24052924]. Hispanics are at greater risk to develop diabetes. Structural barriers limit access to care. Fear among the undocumented, cultural and language barriers contribute to worse clinical outcomes[@Page-Reeves_24052924].

#### Medical insurance status influences health outcomes. 
Medical insurance coverage, which varies by state in the US, predicts access to healthcare delivery and mortality for an individual [@McCarthy_24056035]. As Hoffman reports from a national insurance status survey, patients perceive that the lack or loss of medical insurance covers leads to a plethora of adverse consequences, among them difficulties in paying medical bills, limited access to health services, postponed treatment, failure to obtain medications, especially among low wage workers[@Hoffman_11475546]. Hoffman’s finding echoed similar results by Schoen in an earlier survey[@Schoen_10778809]. McWilliams validated these and other claims in a systematic review of the literature on the effects of insurance status for acute and chronic conditions in adults for which there is effective therapy available[@McWilliams_19523125]. Medical insurance in not enough, but it can make a big difference in health outcomes[@Marmot_24120189], for example LaPar showed that Medicaid and Uninsured payer status confers increased risk-adjusted mortality for major surgery in the national Inpatient Database [@LaPar2010].

## Healthcare Disparities in Surgery, Anesthesia, and Pain Medicine

### Health disparities are recognized in surgery.

#### Trauma outcomes vary according to SES, race and insurance. 
A systematic review by Haider synthesizes the evidence of studies on the influence of insurance status, race and ethnicity, and socioeconomic status on trauma outcomes[@Haider_23609267]. The authors found strong evidence suggesting that insurance status is associated with mortality independent of injury type. Median income predicted higher mortality rates; minority patients have worse outcomes after trauma than their Caucasian counterparts. The effects of race were independent of insurance status. 

#### Insurance status predicts mortality after major surgery
LaPar showed that Medicaid and Uninsured payer status confers increased risk-adjusted mortality for major surgery in the national Inpatient Database [@LaPar2010].

#### Disparities in non-trauma surgical care and outcomes have long been recognized. 
Already in 1977, Egbert described how black patients were several times more likely to receive treatment from a resident surgeon as opposed to a fully trained attending surgeon[@Egbert_325411]. Generations later it is clear that health disparities based on race persist after controlling for insurance status[@Haider_23318117]. Haider led a systematic review of racial outcome disparities in surgery; the authors acknowledged the contribution of provider and systematic factors[@Haider_23318117]. Being a procedure based specialty, racial disparities in the level of training of providers seem important[@Egbert_325411]. This also applies to anesthesia. Resident surgeons involvement can lead to longer case duration at the very least[@Silber_23719571], which in turn can increase infection rates[@Campbell_19183526]. Black patients are at higher risk for poor outcomes after diverticulitis surgery, especially in emergency cases[@Alavi_22125173]. Minorities are more likely to be treated by surgeons with lower surgical volumes, with the latter being linked to poorer outcomes[@Haider_23318117]. The surgical approach may differ with whites more likely to receive less invasive minimal surgery[@Ricciardi_18594921]. 

### Health care disparities are widely acknowledged in pain medicine. 
A systematic review and meta-analysis by Meghani raises alarm about the persistent racial and ethnic disparities in the treatment of pain, clearly a domain of anesthesiologists[@Meghani_22239747]. We described how language barriers may contribute to failure to attend scheduled appointments in chronic pain clinics for underserved inner city populations[@Shaparin_24747766]. Jimenez found racial and ethnic disparities also in perioperative treatment of pain in children, too[@Jimenez_20173265]. 

### Health care disparities likely also exist in anesthesia. 
Healthcare disparities have rarely been investigated in anesthesia to date, however[@Silber_23719571]. Several reports seem to dismiss anesthesia’s contribution to healthcare disparities altogether. Silber reported on the influence of race on operating room time, controlling for obesity and socioeconomic factors[@Silber_23719571]. He found statistically significant and clinically meaningful longer procedure times in black patients[@Silber_23719571]. The authors predicated that the excess in duration was mostly due to longer surgeries as opposed to the anesthesia induction time and emergence; the contribution to the observed disparity by the anesthesia team would hence be limited. After controlling for potential confounders, Toledo no longer found differences in actual use of labor epidural analgesia in a large academic labor anesthesia department[@Toledo_22075013]. Earlier, analyzing the New York State Perinatal Database, Glance in contrast did find lower labor epidural analgesia rates in minorities[@Glance_17197841] . Likewise, Rust described an association between race and ethnicity and epidural rate for labor pain in Georgia medicaid beneficiaries[@Rust_15343221]. Student nurse anesthetists complained about racial discrimination in a survey conducted by Elisha [@Elisha_22403965]. In general, the literature on anesthesia related health disparities seems sparse, apart from labor analgesia and pain medicine. Is there no contribution to healthcare disparity by anesthesiologists or has their contribution just not been established yet?

### Anesthesia is important for improving perioperative outcomes. 
To dismiss the potentially central contribution of anesthesia in observed healthcare disparities is to belittle the impact of anesthesia on perioperative outcomes in general. Glance showed recently that the anesthesia provider had a major impact on mortality after cardiac surgery [@Glance2015]. Anesthesia as a specialty had a significant role in improving surgical mortality over the last generations. Anesthesiologists focused on patient safety starting in the 1970 and as a specialty have been credited with a major reduction in perioperative mortality[@Gaba_10720368;@Leape_7503827]. Silber documented that direction of anesthesia care by highly qualified (board certified) anesthesia providers reduces mortality[@Silber_10861159]. Besides providing anesthesia care in the operating room, anesthesiologists often act as gatekeepers and wardens for the surgical patients before and after surgery and promote an institutional culture of safety[@Gaba_10720368]. As described for surgeons above, this impact may depend on the status and training of the provider. Anesthesiologists with less experience, inferior training or lower status in the hospital hierarchy are likely to have less of a beneficial influence on mortality and morbidity for their patients[@Silber_10861159]. 

## Theoretical Framework: Effectors of health care disparities
Our attempts to counteract healthcare disparities will be most effective if they interrupt the causal chain linking patient characteristics to poor outcomes. It is insufficient to enumerate predictors associated with inferior healthcare outcome. We need to understand the underlying mediators and mechanisms. The models ascribing causality can be classified into three groups:

### Patient and community risk behavior leads to higher mortality and morbidity. 
According to this libertarian perspective, individuals are independent agents endowed with a free will, make poor choices and are therefore responsible for the consequent health risks they incur. Following this logic, patients with lung cancer “chose to smoke,” obese people are responsible for how much they ate[@Knowles1977]. This model is false. Addiction may serve as an obvious counterexample. Addiction is a disease[@McLellan_11015800] . It does not make sense to argue that patients choose to be addicted. Page-Reeves deconstruct the “free will” model using the example of diabetes care in New Mexico[@Page-Reeves_24052924] . Page-Reeves instead argues for a model considering social determinants of health and health relevant behavior[@Page-Reeves_24052924] . The concept of health behavior choices being expressions of deliberate free will has repeatedly been challenged; for example Lynch showed that health behavior patterns may be imprinted in earlier life stages[@Lynch_9080564]. Indeed, childhood parental socioeconomic status explaining much of adult risk behavior[@Lynch_9080564]. The Institute of Medicine report also unequivocally showed that the health disparities associated with race are not explained by inability to pay, lack of insurance or even attitudes or treatment decisions by African American patients; rather, these disparities are the result of systematic racial discrimination by their physicians and providers[@IOM_2003]. 

### Discrimination at the provider level may explain inferior outcomes. 
Subconscious or intentional bias may lead physicians and other providers to administer differential care according to race or ethnicity contributing to inferior health outcomes[@Santry_22269267]. The US Congress tasked the Institute of Medicine with a report on racial and ethnic healthcare disparities. Nelson summarized how stereotyping, bias and outright discrimination by providers leads to inferior health outcomes in the US[@IOM_2003;@Nelson_12152921]. These differences in the quality of medical care delivered to ethnic and racial minorities were not explained by barrier to access care, patients’ attitudes & decisions or their inability to pay. 

### Systematic bias in the healthcare delivery system results in lower health status. 
Health care system bias could also be impersonal, expressed in the rules and regulations or unequal geographic allocation of resources[@McCarthy_24056035]; they can unfairly lead to inferior treatment options for minorities or stigmatized populations, also detailed for example by Nelson[@Nelson_12152921]. A good example of the impersonal systematic bias is the inferior care afforded to immigrant and undocumented populations[@Page-Reeves_24052924]. Page-Reeves illustrates how regulatory forces can significantly hamper access to care and cause insecurity among immigrant populations with resultant poorer health outcomes[@Page-Reeves_24052924]. Page-Reeves goes so far as to calls this “structural violence”[@Page-Reeves_24052924]. 

## Finding the Culprit: Who is Responsible for Healthcare Disparities?

#### This paper seeks to investigate if disparities exist also in anesthesia. 
More precisely, we seek to establish if the quality of anesthesia care differs depending on the medical insurance status. We seek to investigate healthcare disparities at the provider level. On a personal note, the author enjoys working in anesthesia not least because anesthesiologists typically need not care about the insurance status of their patients. If a patient makes it to the operating room, insurance questions have either been resolved or are suspended for emergency surgery. Anesthesiologists operate under the pretense that we treat all-comers equal. Yet anesthesiologists are not free of racial bias or other prejudice. We simply ignore if these may lead to disparate anesthesia care unbeknownst to us and/or against our best intentions.  

#### Identifying causal factors of healthcare disparities is difficult. 
Differentiating which actors or factors in the health care delivery process are responsible for observed healthcare disparities can be perplexing, especially when it comes to anesthesia and perioperative care[@Silber_23719571] . We are facing several challenges. 

1. We seek to single out the effect of the anesthesia provider from among the contribution of all actors in the perioperative process. 
2. We need to control for the many potential confounders of the relationship between medical insurance status and healthcare disparities. 
3. We would like to ensure that the observed detrimental effect or possibly the inferior care delivered is unlikely to be due to or influenced by monetary considerations. Either incentives to the providers or restrictions on the part of the payer are examples of systemic impersonal mechanisms causing healthcare disparities.

#### Many actors are involved in the perioperative process.  
Refereeing physicians, admitting personal, surgeons, anesthesiologists, nurses in the postoperative care unit and on the floor and after hospital care facilities all have their role in providing comprehensive surgical care. Silber, in his investigation of racial disparities in operative procedure time, observed longer durations in surgeries of black people. Most of the difference was accounted for by the time between surgical incision and wound closure, i.e. the operation proper. This suggested that the surgical team operating was responsible for any difference in duration (Silber, et al., 2013). The contrary is difficult to prove. Anesthesiologists could however have contributes to the excess duration of the procedure time observed in Blacks. We will list just a few examples from cardiovascular /general surgery and orthopedics. (a) An insufficiently fluid pre-loaded patient may need several attempts during unclamping of the aorta to stabilize before surgery can carry on. This can be the fault of poor choices by the anesthesia provider. (b) If the airway is insufficiently secured, the surgeon may be forced to halt a procedure until it is safe to proceed. If the stomach is not deflated (by the oro-gastric tube inserted by anesthesia) poor visualization or access can hinder surgical progress. (c) An insufficient regional anesthetic block, inadequate neuromuscular relaxation or inadequate sedation light anesthesia may slow surgery if the patient continues to move about under the scalpel. We conclude that Silber’s study is inconclusive as to how much the surgical versus the anesthesia team contributed to the observed difference in length of procedure time and hence to healthcare disparities (Silber, et al., 2013). 

#### Many patient factors predict healthcare disparities. 
Some of these are highly co-linear. For example race and socioeconomic status are statistically so closely associated that including one in a regression model often removes any apparent contribution of the other. It is often very difficult to separate their effects even with advanced statistical methods(Cook, McGuire, & Zaslavsky, 2012). We need to control for geographic variations in care on the state and down to the county level (Cooper, et al., 2012), if we examine national registries like the NACOR we propose to study. 

#### Difference in insurance may dictate differential treatment. 
The hands of a provider are tied if the insurance limits the therapeutic options or if they are restricted by hospital administrators ordering certain clinical pathways. An example might be the choice of neuromuscular blocker used, which can affect postoperative complications like aspiration pneumonia or prolonged ventilation increasing the risk of ventilator associated pneumonia. Administrators may restrict choices neuromuscular blocking agents as some (e.g. rocuronium) account for a large share of the hospitals pharmaceutical budget. Hospital administrators on the other hand seek to maximize payments by providers by ensuring documentation of certain aspects of care considered markers of quality. Examples include administration of beta-blockers and application or warming blankets. The actual treatment administered may differ from the documentation (a) the more high compliance is enforced by the billing department and (b) the less providers accept the evidence supporting such therapy. If we seek to identify health care disparities at the individual provider level, we should hence try to observe documented aspects of care not influenced my monetary constraints or regulatory reporting requirements. 

## Antiemetic utilization as a surrogate marker of quality of anesthesia care

A suitable surrogate marker for the quality of anesthesia care delivered at the provider level would hence have three attributes: 

1. The marker is independent of patient characteristics (comorbidity, genetic disease, health status, ability to pay, patient attitudes and choices) and confounders (income, education level, geographic location). 
2. Administration of the marker is the sole responsibility of the anesthesia provider without input by any other actors in the perioperative process (surgeon, nurses, admission, pathology). 
3. The utilization of the marker is unlikely to be influenced by regulatory, monetary or insurance constraints or considerations (reporting requirements, cost, and financial bonus or billing aspects). We propose to investigate antiemetic prophylaxis as a surrogate marker of anesthesia quality, because we feel this marker has all three above postulated attributes.

#### Postoperative nausea and vomiting is an important outcome. 
Besides the individual unpleasant experience by the nauseated patient, PONY is one of the main reasons of prolonged stay in the recovery unit and hence prophylactic antiemetic interventions have widespread impact in perioperative care of the patient[@Apfel_15190136]. Every diligent anesthesia provider should consider antiemetic prophylaxis without consideration of insurance status. Antiemetic administration is especially indicated in patients at high risk for PONY (females, young patients, non-smokers, and history of PONY or motion sickness)[@Apfel_15190136]. There are two principle choices of antiemetic prophylaxis, ondansetron and dexamethason. On one hand, ondansetron is still under patent protection and therefore costly; its administration may hence be constraint by monetary considerations. Also ondansetron is contraindicated in rare certain cardiac conditions (QT prolongation). On the other hand, dexamethason is off patent, widely available, extremely cheap and without any contraindications. Either of the two antiemetic agents should be available and suitable for prophylaxis of postoperative nausea and vomiting (PONY) in any given patient.
We assume that utilization of antiemetic prophylaxis is a suitable surrogate marker of quality of anesthesia care. We consider medical insurance as a surrogate marker for socioeconomic status. We want to investigate the effect of socioeconomic status on quality of anesthesia provided by the individual anesthetists in the subset of the National Anesthesia Outcome Registry (NACOR) with full electronic anesthesia records[@Dutton_21871404]. 

## Hypothesis

**Medical insurance status predicts antiemetic prophylaxis during surgery as a surrogate marker of quality anesthesia care.** 

Contingent on our assumption, if our hypothesis is true, American anesthesia providers on average may provide different quality of anesthesia depending on the insurance status of their patients. Inferior anesthesia quality may lead to inferior perioperative outcomes[@Santry_22269267;@Silber_23719571]. American anesthesiologists as a specialty would hopefully accept the responsibility for the described health care disparities in anesthesia and take corrective action to ensure equitable high quality care to all their patients. 


-----

\newpage

# Methods

## Description of the sample studied 

### The National Anesthesia Clinical Outcomes Registry
The National Anesthesia Clinical Outcomes Registry (NACOR) receives information on anesthesia cases from participating institutions and anesthesia providers[@Dutton_21871404]. The subset of NACOR containing complete electronic anesthesia records (including dose and time of antiemetic prophylaxis administered) comprises about one million anesthesia cases. We to studied this dataset. Datasets contain about 65 variables for each case. Variables span information on the provider of the anesthesia (ID, location, class (university hospital, private anesthesia group)), on the patient (Age, gender, anesthesia risk classification, insurance status, disease) and on the procedure (Billing code, modifiers, indication ICD code).  The data is heterogeneous with many missing data, as not all participating participants upload the same information. Some participating providers may only upload billing data, other the full electronic anesthesia record containing even individual physiological parameters recorded minute by minute. 

### Or custom enriched dataset
We used data downloaded from the Public User File (PUF) of the National Anesthesia Clinical Outcomes Registry, from Quarter 4 of 2014, provided and  enriched with additional information on antiemetic useage and insurance status by the Anesthesia Quality Institute. The data had been ubloaded by participating provider institutions. There is a minimum dataset a participating provider can upload containing mostly demographics, but only a small portion of all providers upload the complete electronic anesthesia record. This dataset mostly contains intra-operative vital data and administered medications, but little if any postoperative outcome data after anesthesia.

We described the characteristic of the subset of NACOR datasets forming the bases of our analysis, anesthesia records with complete information on the administration of antiemetic prophylaxis. We explored the bivariate associations between the dichotomous  outcome variable antiemetic prophylaxis and the independent variables describing patients, procedures & providers; patient characteristics included medical insurance status (our primary predictor of interest), patient age, gender, American Association of Anesthesiology risk classification, co-morbidities, (but not race, which is not recorded in NACOR). We reported procedure types and indications (Billing code, modifiers, indication ICD code). Provider characteristics include information on the anesthetist [nurse anesthetist versus resident versus attending alone] and institutional data [geographic location, academic versus private versus government institution]. 

We used parametric tests where the assumptions of normality did not seem violated and non-parametric test, where graphical or statistical tests suggested possible violations of the underlying assumptions. We reported proportions, mean and standard deviation or the median and the interquartile range (as appropriate for the distribution of values observed for each parameter) and indicated the statistical test used in the table of characteristic of patients. 

## Statistical analysis

We fit a Bayesian and a classical  logistic regression in the subset of anesthesia cases in the National Anesthesia Clinical Outcomes Registry (NACOR) with information on medical insurance status and intra-operative antiemetic administration. Administration of antiemetic prophylaxis is the primary outcome and medical insurance status is the ordinal predictor, controlling for patient & provider characteristics. We will used three different approaches to analyse the data:

1. a logistic regression model, 
2. a multivariate stratification and
3. a hierarchical Bayesian model.  

## Software used

we used the statistical software Stata\textsuperscript{\textregistered} for the regression analysis[@Stata_2014]. We used R, the public domain statistical software package[@R_software] and the probabilist programming Software Stan[@Stan_Software_2014] to implement the hierarchical Bayesian model. R and Stan are available under the General Public License of the Free Software Foundation[@GNU_1991] at no cost. We used the software package shinyStan[@shinystan-software_2015] for graphical exploration of model convergence, to generate the contrasts to compare commerical versus medicaid and for posterior predicitive checking.

## Modeling

We investigated the effect of medical insurance as predictor on the administration of antiemetic medication as primary outcome, controlling for potential confounders like patient characteristics, provider characteristics and procedure type and indication. The predictor insurance status is an ordinal variable; possible values are ordered from highest insurance coverage, i.e. private insurance, Health Maintenance Organization, medicare, medicaid, to the lowest no medical insurance reported. Our outcome is dichotomous, antiemetic prophylaxis administered or not. Our unit of analysis is the anesthesia case, not the patient. We focused our analysis on the most frequent procedures performed. We considered findings statistically significant if the p-value was less than the alpha of 0.05.Patients may have several operations, each generating one anesthesia record; patients may hence be counted several times. We will only consider antiemetic prophylaxis administered during anesthesia care, either in the operating room or while dropping off the patient in the recovery unit, but before sign-out to the recovery staff. 

### Classical logistic regression
We fit a classical logistic regression model. Besides insurance status as our primary outcome, we decided a priori to include gender and age in model, because both have been previously shown to be risk factors for PONY and are hence considered indications for antiemetic prophylaxis; as such they may act as confounders of associations. In addition, we  chose those independent variables for the initial model that show a statistically significant association in the bivariate analysis. We used stepwise backward elimination starting from our initial model based on the likelihood ratio test with a cutoff at 0.05 to eliminate independent variables from the model. For each model eliminated we confirmed that the given variable was not a confounder for the present model. We used a change in the beta coefficient of larger than 20% as our cutoff to determine if a variable is considered a confounder. We determined the correct functional form and explore potential violations of the assumptions of linearity. We ran locally weighted regression of yvar on xvar (and examined the graph for all independent variables in our final model), for a graphical assessment of potential violations of the assumption of linearity. We tested for the correct functional form, fitting fractional polynomials as part of our final logistic regression model. We  examined if the addition of a polynomial improves the model significantly. We explored the potential interaction between the independent variables age and gender in a simple logistic regression model; a cut of for our likelihood ratios test was at an alpha level of 0.05. We examined the goodness of fit with the Hosmer-Lemeshow goodness of fit test. We performed a sensitivity analysis of our model assumptions and choices.

### Stratified analysis
We calculated odds ratios for the association between insurance status and antiemetic administarion with the data stratified by gender, age and other demographics and case charactersitics.
 

### Bayesian hierarchical model
We build a hierarchical Bayesian model. We present the [Bayesian Model Code] in the appendix. We modeled the outcome as a dichtomous event $y_a$ for each anesthetic given, our unit of analysis

$$
  \begin{aligned} 
  y_{a} &\sim& Bernoulli(\mu_{a})\\
  \mu_a &\sim& inverse\_logit(\gamma[g_a] + x_a*beta)))
   \end{aligned}
$$

where $\mu_a$ is the probability of an event (ondansetron administration) in the a^th^ anesthetic, $g_a$ indicates the institution (where the anesthetic was given to the patient), $\gamma$ is the random effect each institution has on the administration of the antiemetic in a given case, $\beta$ the the coefficient of effect for the contrast between medicaid and commerical insurance and x_a the indicator for the insurance status.

Cases are nested in providers and these will be nested in institutions. (There are no patient identifiers to track repeated operations to the same patient.) We used uninformative priors for the main effect of insurance status on antiemetic prophylaxis. We performed a sensitivity analysis investigating the appropriateness of our model assumptions. 

We compared the results of our three models (logistic regression, stratified analysis and Bayesian hierarchical model) to confirm the robustness of our findings regardless of the choice of model or the statistical approach chosen.

-----

\newpage

# Results

## Description of the dataset

Our analysis is focused on the subset of the above subpopulation with information on both antiemetic usage *and* insurance status. The prevalence of ondansetron administration for different insurance status is  described in [Table 2 Insurance status versus antiemetic prophylaxis]. Clearly,  Patients who have commerical insurance are more likely to receive antiemetic prophylaxis than those who have medicaid.

Unfortunately, only 441645 cases in the Public User File (PUF) contained detailed information on medications administered during anesthesia administration at the end of 4th quarter 2014. Six unique institutions reported insurance status and antiemetic utilization for 176,334 anesthesia cases, out of the 12 million cases in NACOR 2014. Limiting this set further to cases with information on medical insurance, the dataset of our study contained 176334 anesthesia cases.The reporting institutions were mostly university hospitals or medium to large community hospitals, mostly in the Southern United States . Our dataset contained no cases from the Midwest or the West of the Unites States. Anesthesia was provided between 2010 and 2014. 

### Population
The demograpics of the subpopulation in the NACOR database with information on antiemetic administration are described in [Table 1 Population characteristics]. 42 percent of anesthetics were administered to male patients. Patients’ age ranged from newborn to 90 years of age with a median of 54 (Interquartile range (IQR) between 35 and 67 years). Most patients were classified as ASA class 2, 3 or 4 (39, 37 and 16 %, respectively) with only 7 percent in class 1 or 5.  62 percent were outpatients among the 64% of cases where this information was available. For 25865 cases the insurances was reported as medicaid, for another 97443 as medicare for the remaining 51441cases mostly as commercial insurance with 1585 self-insured cases. Some antiemetic was administered in 55 percent of all cases. 

### Bivariate comparison of demographic charactersitics
We found  statistically significant differences (p<0.001) in all demographic bivariate comparisons; they suggests that patients who received ondansetron (Zofran\textsuperscript{\textregistered}) were on average younger and more likely to be female, to have a slighly longer procedure, but as an outpatient, less likely to have undergone emergency surgery, more likely to live in the South in a zip code with higher mean income and smaller population size. 

## Association between insurance status and antiemetic administration

### Logistic regression analysis
We show the results of our final logistic regression model in the [Table 3 Logistic regression]. Being on medicaid or medicare, compared to having commerical insurance, drastically reduces the odds of receiving ondansetron during anesthesia. For the average patient on medicaid or medicare, the odds of receiving ondansetron for anti-emetic prophylaxis are 0.6 compared to a patient with commerical insurance with otherwise similar characteristics, even after controlling for age and gender, case duration, median income and population in the patient's home zip code. 

### Stratified analysis

We performed a stratified analysis, calculating the odds ratios for receiving ondansetron for patients on medicaid insurance versus commercial insurance. We show the results in the [Table 4 startified odds ratios] for the stratification by gender, older age, general anesthesia, outpatient surgery or the presence of a board certified anesthesiologist.  Stratification did change the odds ratios. However in all strata,  medicaid insurance status was associated with reduced odds of receiving antiemetic prophylaxis suggesting that the strong assoication between insurance status and antiemetic prophylaxis holds accross the board.

### Hierachical Bayesian model

[Table 5  Bayesian model] shows the parameter estimates for our simple Bayesian hierarchical model in a table. This Bayesian model only controlled for the effect of the institution as a random effect. Also in this model, having medicaid versus commerical insurance  was associated with receiving anti-emetic prophylaxis with an odds ratio of 0.6 with a narrow credible 95% interval [0.59, 0.62] . Remarkably, the random effects estimated for the different institutions in which the surgery took place vary considerably, suggesting that healthcare disparities may be endemic, or locally more or less pronounced.

#### Our Bayesian model converged. 
The convergence of our Bayesian model was confirmed by looking at traceplots and the Gelman Rubin statistics[@R_hat_Gelman_1992], shown for selected parameters of our Bayesian model.

## Sensitivity Analysis

The strong and statistically significant association (p < 0.001) between insurance status and the odds of receiving antiemetic prophylaxis remained unchanged after stratification to control for patient characteristics like gender and age. Our inferences were invariant to the statistical approach (Bayesian versus classical frequentist analysis) we used and bore out in the subset with data for multivariate logistic regression as well as in stratified analysis.  

# Discussion

## Summary of the findings

Bivariate analysis, multivariate logistic regression and our Bayesian hierarchical model all showed a strong and statistically significant association between insurance status and the odds of receiving antiemetic prophylaxis; for medicaid versus commercially insured patients, the odds ratio of receiving the antiemetic ondansetron is 0.63 in our final logistic regression model with a 95% confidence interval of [0.62, 0.66] with similar Bayesian inferences. 

## Strength
The size of the subset of NACOR we studied likely makes this the largest study of healthcare disparities in anesthesia undertaken to date. This increased our power to detect an association between insurance status and antiemetic prophylaxis. Controlling for likely confounders decreases the chance that the association is spurious.

## Impact
Demonstrating health care disparities in such a large dataset in anesthesia is novel. The fact that the surrogate marker of anesthesia quality is exclusively in the domain of the anesthesia will likely make a greater impression on the anesthesia community than an intervention or marker for which anesthesiologists are only partially involved or responsible for, like procedure time [@Silber_23719571]. Our work clearly puts the onus on the individual provider as opposed to system bias due to regulations or geographic variations. Our findings may hopefully lead to a somber assessment of healthcare disparities in anesthesia by all stakeholders, and appropriate counter measures. The anesthesia community has risen to the opportunity in the past and provided leadership in provision of equitable and safe care to all our patients[@Gaba_10720368].

## Limitations

### External and internal validity of our findings 

#### Our population may not be representative
Our sample is likely not representative of the general patient population undergoing anesthesia in the US, but oversampled academic institutions and the North East; we only have data from the institutions that uploaded full electronic anesthesia records; these institutions and their anesthesia practices are likely different from the average practice, e.g. academic practice is very different from private practice, with the later presumably catering more to outcomes patients perceive as important, PONY being one of them. 

#### Selection bias may limit our conclusions. 
Not all providers and institution use electronic anesthesia records and if they do they may not upload the data to NACOR, e.g. for regulatory or privacy reasons. Will certain providers with greater likelihood to discriminate on insurance status be less likely to have electronic medical records or to upload their datasets? This possibility may limit the generalizability of our findings as this may result in selection bias. Hence, our analysis should be repeated with a broader sample  as soon as additional patients data are uploaded to NACOR.

#### Information bias may result in under- or overestimation of the observed effect. 
Providers may have forgotten to record the administration of antiemetic medication, but still have administered the prophylaxis. They may have recorded the dose, but might still have failed to administer the prophylaxis. Both instances of misclassification of outcome (antiemetic prophylaxis administration) could lead to an over- or underestimation of healthcare disparities as we defined them. This distortion of the effect would be larger if the misclassification were not completely at random but differential depending on our predictor of interest (health insurance status). Because antiemetic prophylaxis is generally not considered critical or relevant for billing and malpractice, malicious intent to falsify the record is unlikely to lead to numerous differential misclassifications. Sloppy failure to record the medication, while actually giving it, seems a more likely occurrence. However, poor record keeping is likely associated with poor anesthesia quality and hence this misclassification would not invalidate the conclusions we draw from our analysis.

#### Our conclusions hinge on our assumption. 
If antiemetic prophylaxis is not a good indicator of anesthesia quality, then we cannot conclude that healthcare disparities exist in anesthesia, even if we find that insurance status predicts antiemetic administration. Likewise, if or as soon as antiemetic prophylaxis is used widely as such a marker, bias will be introduced in the documentation and administration of these agents and they may no longer genuinely reflect quality. Analogously, insurance status while closely linked to race and socioeconomic status is only one predictor of healthcare disparities and we do not know what might cause providers to differentiate based on the former versus the latter and/or if the same causal pathways are involved.

### Sensitivity analysis   

#### Each of our statistical approaches has its limitations. 
Stratified analysis only controls for the specific variable by which the stratification is done and doing substratification quickly becomes overwhelming. Logistic regression does not control well for confounding when the probability of effect is around 0.5. Hierarchical models lead to shrinkage, which makes for a better fit of the model to the data, but this may lead, among other issues, to overfitting.

#### Our results were invariant to the statistical approach.
If we used stratification, hierarchical Bayesian modeling or classical frequentist logistic regression, we came to the same inferences. But while we controlled for some important patient characteristics, we still need to consider surgery and provider as confounders and implement a full Bayesian hierarchical model. Certain surgical interventions are (believed) to be associated with a higher incidence of PONY and in some,is very undesiable, (e.g. because  the wretching concomittant with PONY can put undue stress on the surgical closure.) This can lead to confounding, if providers selectively administer antiemetics in these cases and surgical intervention and insurance status are associated. 


## Continous improvement is the ideal in healthcare
In the process of qulity improvement, we typically go through the four stages described by Don Berwick[@Berwick_2909878]:

1. Stage I: “The data are wrong….”
2. Stage II: “The data are right, but it's not a problem…”
3. Stage III: “The data are right, it's a problem, but it's not my problem…”
4. Stage IV: “The data are right, it's a problem, it's my problem…”

We hope to convince anesthesiologists with the presented strong and consistent association between insurance status and antiemetic prophylaxis regardless of the statistical approach, that the findings are solid. We think there is a clear argument that the described assoicatation suggest healthcare disparity for which anesthesia providers are accountable. The next step is to work to convince the community that this is our problem and that we can and should do something about this. Continous improvement is the ideal in healthcare[@Berwick_2909878] and we as anesthesiologist have a tradition to defend to be the leaders in continued improvement addressing individual performance as well as systems to achieve the best of care for all of our patients[@Gaba_10720368]

# Conclusion: 

Our analysis of the association between insurance status (as a marker of patient socioecomonic status) and antiemetic administration (as a marker of anesthesia quality provided by the individual anesthesiologist) point to possible unappreciated healthcare disparities in anesthesia at the provider level which are worrisome and deserve further investigation and vigorous countermeasures.


\newpage

# Tables

## Table 1 Population characteristics {#Table1}

```{r,eval=FALSE}
# C:\Users\Michael Andreae\Dropbox\Professional\AQI\Robert\tables for abstract
library(knitr)
Table <- read.csv("C:/Users/Michael Andreae/Dropbox/Professional/AQI/Yeti/AQI Yeti/Table.csv")
kable(Table[1:9,1:4], row.names=FALSE)
```

Table 1 Population characteristics describes the subpopulation in the NACOR database with information on antiemetic administration. We found  statistically significant differences (p<0.001) in all comparisons (age, gender, income in the patients zip code and case characteristics like outpatient procedure, emergency procedure and case duration.

## Table 2 Insurance status versus antiemetic prophylaxis{#Table2}

```{r,eval=FALSE}
kable(Table[10:14,1:4], row.names=FALSE)
```

Table 2 describes the subset of patients (among the subpopulation from Table 1 Population characteristics with information on antiemetic administation) who have data on insurance status. Clearly  Patients who have commerical insurance are more likely to receive antiemetic prophylaxis than those who have medicaid.

## Table 3 Logistic regression {#Logistic_table}

```{r,eval=FALSE}
# AQI_stata_data_analysis.Rmd
rm(list=ls())
require(knitr)
load("table_my_logistic2.Rdata")
kable(table_my_logistic2, digit=3, row.names=TRUE)
```

The results of our logistic regression, presented in a table as odds ratios with the coresponding 95% confidence intervals, show that being on medicaid or medicare, compared to having commerical insurance, drastically reduces the odds of receiving ondansetron during anesthesia.

## Table 4 startified odds ratios 

```{r}
rm(list=ls())
require(knitr)
load("Tables/LOR_table.Rdata")
kable(OR_table, digits=3)
```

This table shows the odds ratios for receiving ondansetron for patients on medicaid versus patients on commercial insurance after we stratified for various possible confounders. Stratifying for gender, age, general anesthesia, outpatient surgery or the presence of a board certified anesthesiologist did change the odds ratio. However in all strata,  medicaid insurance status was associated with reduced odds of receiving antiemetic prophylaxis suggesting that the association holds accross the board.


##Table 5  Bayesian model {#Bayesian_table}

```{r}
rm(list=ls())
require(knitr)
load("Tables/Bayesian_table.Rdata")
kable(Bayesian_table[,c(2,1,3,4,5)], digits=3 )
```

Having medicaid versus commerical insurance reduces the odds of receiving anti-emetic prophylaxis significantly in our Bayesian hierarchical model. This is clear from the estimated odds ratio of receiving ondansetron for the contrast of $\beta_1$ (medicaid) versus $\beta_3$ (commercial) shown with the corresponding 95% Bayesian credible intervals. Remarkably, the random effects estimated for the different institutions in which the surgery took place vary considerably, suggesting that healthcare disparities may be endemic, or locally more or less pronounced.

\newpage

# Appendix

## Bayesian Model Code

~~~~~~~

functions {
  
  /** 
   * Transform standard normal to cauchy(location, scale).
   *
   * @param location Location parameter for Cauchy distribution
   * @param scale Scale parameter for Cauchy distribution
   * @param noise Standard normal variable (as declared in parameters block)
   */
  real cauchy_trans_lp(real location, real scale, real noise) {
    noise ~ normal(0,1) ;
    return location + scale * tan(pi() * (Phi_approx(noise) - 0.5)) ;
  }
  
  /** 
   * Transform standard normal to normal(location, scale).
   *
   * @param location Location parameter for normal distribution
   * @param scale Scale parameter for normal distribution
   * @param noise Standard normal variable (as declared in parameters block)
   */
  real normal_trans_lp(real location, real scale, real noise) {
    noise ~ normal(0,1) ;
    return location + scale * noise ;
  }
  
  /** 
   * Transform vector of standard normals to vector of normals
   * with same location (mean) and scale (sd).
   *
   * @param location Location for normal distributions
   * @param scale Scale for normal distributions
   * @param noise Name of vector of standard normals (as declared in parameters block)
   */
  vector normal_trans_vec_lp(real location, real scale, vector noise) {
    noise ~ normal(0,1) ;
    return location + scale * noise ;
  } 
}

data {
  // dimensions of data
  int<lower=1>                  nOb ;     // number of cases
  int<lower=1,upper=nOb>        nGrp ;    // number of facilities
  int<lower=1>                  nX ;      // number of predictors
  
  // data 
  int<lower=0,upper=1>          y[nOb] ;   // outcome (antiemetic administration) 
  int<lower=1,upper=nGrp>       g[nOb] ;   // map obs --> groups (cases --> facilities)
  row_vector[nX]                x[nOb] ;   // predictors (dummies for insurances status)

  // locations and scales for priors
  real                          beta_loc ;        // beta = fixed effects for predictors
  real                          sigma_gamma_loc ; // sigma_gamma = scale for gamma
  real                          gamma_loc ;       // gamma = random effects for groups
  real<lower=0>                 beta_scale ;
  real<lower=0>                 sigma_gamma_scale ;
}

parameters {
  vector[nX]        beta_noise ;    
  real<lower=0>     sigma_gamma_noise ;   
  vector[nGrp]      gamma_noise ;   
}

transformed parameters {
  vector[nX]        beta ;          // fixed effects for predictors
  real<lower=0>     sigma_gamma ;   // variability (for gammas)
  vector[nGrp]      gamma ;         // random effects for groups
  
  beta        <- normal_trans_vec_lp(beta_loc, beta_scale, beta_noise) ;
  sigma_gamma <- cauchy_trans_lp(sigma_gamma_loc, sigma_gamma_scale, sigma_gamma_noise) ;
  gamma       <- normal_trans_vec_lp(gamma_loc, sigma_gamma, gamma_noise) ;

  /* 
  Priors implied by transformations:
      beta[i] ~ normal(beta_loc, beta_scale)
      sigma_gamma ~ half-cauchy(sigma_gamma_loc, sigma_gamma_scale)
      gamma[j] ~ normal(gamma_loc, sigma_gamma) 
  */
}

model {
  vector[nOb] mu_logit ; // linear predictor

  for(n in 1:nOb) 
    mu_logit[n] <- gamma[g[n]] + x[n]*beta ;
  
  y ~ bernoulli_logit(mu_logit) ; // vectorized 
}

/*
generated quantities {
  int y_rep[nOb] ;  // replicated data for posterior predictive checking
  
  for (n in 1:nOb)
    y_rep[n] <- bernoulli_rng(inv_logit(gamma[g[n]] + x[n]*beta)) ;
}
*/

~~~~~~~

\setlength{\parindent}{0ex}


# References