---
title: \normalsize Utilization of Antiemetic Medication as a Marker of Healthcare
  Disparities in Anesthesia \newline A Bayesian Hierarchical Model Using the National
  Anesthesia Clinical Outcomes Registry
csl: Bibliography/anesthesia-and-analgesia.csl
fontfamily: pslatex
output:
  word_document: 
    reference_docx: mystyles.docx
bibliography: Bibliography/AQI_bibliography6May2015.bib
---

\setlength{\parindent}{0ex}

# Authors

Michael H. Andreae, MD, Department of Anesthesiology, Montefiore Medical Center, Albert Einstein College of Medicine, Bronx, NY

Jonah Galbry, MSc, Department of Statistics, Columbia University, New York, NY

Robert S White, MD, Department of Anesthesiology, Weill Cornell Medical Center, New York, NY

Charles Hall, PhD, Department of Epidemiology & Population Health, Albert Einstein College of Medicine, Bronx, NY

# Corresponding author:

Michael Andreae, MD, Address: Department of Anesthesiology, Montefiore Medical Center, 111 E 210th Street, Bronx, NY 10467, Tel: +1 (718)920 2802, Email: mandreae@montefiore.org

# Acknowledgement

This research is supported in part by the National Center for Advancing Translational Sciences (NCATS), a component of the National Institutes of Health (NIH), through CTSA grant numbers 5KL2TR001071-03 and UL1 TR001073. Its contents are solely the responsibility of the authors and do not necessarily represent the official views of the NIH.

**Number of words in Abstract (400), in Introduction (500), and in Discussion (1,500)**

\newpage

#Abstract

**Background:** US Health care disparities persist despite repeated countermeasures. Research identified race, ethnicity, gender, and socio-economic status as predictors, mediated through individual provider and/or systemic biases; little research exists in anesthesiology. We investigated antiemetic prophylaxis as a surrogate marker for anesthesia quality by individual providers, because antiemetics are universally available, indicated independent of patient co-morbidities and not yet impacted by regulatory or financial constraints. 

**Methods:** We hypothesize that socioeconomic markers (like insurance status or median income in the patients home zip code area) predict antiemetic prophylaxis as a marker of anesthesia quality. We fit a mixed logistic regression models in the subset of anesthesia cases in the National Anesthesia Clinical Outcomes Registry (NACOR) with complete electronic anesthesia records to test our hypothesis. 

**Results:** Seven institutions reported insurance status and antiemetic utilization for 176,334 anesthesia cases, out of the 12 million cases in NACOR 2014. Bivariate analysis, multivariate logistic regression and our Bayesian hierarchical model all showed a strong and statistically significant association between socioeconomic markers and the odds of receiving antiemetic prophylaxis; for medicaid versus commercially insured patients, the odds ratio of receiving the antiemetic ondansetron is 0.63 in our final logistic regression model with a 95% confidence interval of [0.62, 0.66] with similar Bayesian inferences. 

**Discussion:** Our data support the notion that individual anesthesia providers administer lesser quality care to patients with lower socioeconomic status. We controlled for some important patient characteristics as well as for procedure and provider. Our sample may not be representative and may fail to account for all confounders. Our results point to possible unappreciated healthcare disparities in anesthesia at the provider level which are worrisome and deserve further investigation and vigorous countermeasures.

```{r, global_options, echo=FALSE}
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

\newpage

\setcounter{secnumdepth}{2}
\setlength{\parindent}{10ex}

# Introduction (500 word limit)

## Background
The healthcare disparities in the United States of America described decades ago by Gornick[@Gornick_8703185], persist and are linked to social determinants of health and equality[@IOM_2003;@Cooper_22566148; @Schoen_24226092].
Poverty, poor education, differences in medical insurance coverage, geographic location, legal or social status, race & gender, patient and community attitudes & perceptions and, last but not least, provider bias lead to healthcare disparities[@Silber_23719571]. A systematic review by Haider suggested that insurance status, median income, race, ethnicity, and socioeconomic status predict trauma outcomes, independent of injury type[@Haider_23609267]. LaPar showed that Medicaid and uninsured payer status confer increased risk-adjusted mortality for major surgery in the National Inpatient Database [@LaPar2010].

## Significance
Do healthcare disparities exist also in anesthesia? A systematic review and meta-analysis by Meghani raises alarm about the persistent racial and ethnic disparities in the treatment of pain, clearly a domain of anesthesiologists[@Meghani_22239747]. We described language as barrier to access chronic pain services[@Shaparin_24747766]. Jimenez found disparities in pain treatment in children[@Jimenez_20173265]. Unfortunately, apart from labor analgesia[@Toledo_22075013; @Glance_17197841; @Rust_15343221; @Elisha_22403965] and pain medicine[@Glance2015;@Gaba_10720368;@Leape_7503827;@Silber_10861159], the literature on anesthesia related health disparities seems sparse[@Silber_23719571]. 

## Objective

We propose to investigate antiemetic utilization (AU) as a marker of quality anesthesia care and health care disparity. Three characteristics make AU a good marker: 

1. Antiemetic utilization is relatively independent of patient characteristics (e.g. genetic disease, ability to pay or geographic location), 
2. antiemetic utilization is the sole esponsibility of anesthesia providers,
3. antiemetic utilization is not influenced by regulatory, monetary or insurance constraints and has a an accepted standard of care [@Apfel_15190136].

This paper investigates if antiemetic utilization is predicted by socioeconomic status. We explore the association of the two predictors insurance status and median income in the patients' home zip code with antiemetic utilization of ondansetron and/or dexamethason in the subset of the National Anesthesia Clinical Outcomes Registry (NACOR) with full electronic anesthesia records [@Dutton_21871404].

## Hypothesis

Our hypothesis is that socioeconomic patient characteristics will predict antiemetic utilization after controlling for institutional and geographic variability, patient preoperative risk factors, procedure and anesthesia type and controlling for individual provider behavior.

\newpage

# Methods
We obtained the Public User File (PUF) of the National Anesthesia Clinical Outcomes Registry (NACOR), from Quarter 4 of 2013, provided and enriched with additional information on antiemetic useage and insurance status by the Anesthesia Quality Institute (AQI). AQI had removed all patient identifiers.

The Albert Einstein College of Medicine Institutional Review Board determined that in accordance with the OHRP Guidance on Research Involving Coded Private Information or Biological Specimens, our study does not meet the definition of human subject research as defined by 45 CFR 46.102(f), as the data/specimens were not collected specifically for the proposed research project and are completely de-identifed.

NACOR receives information on anesthesia cases from participating institutions and anesthesia providers[@Dutton_21871404]. The data had been ubloaded by participating provider institutions. There is a minimum dataset participating provider upload to NACOR, containing mostly demographics. Only a small portion of providers aditionally upload the complete electronic anesthesia record including intra-operative physiologic data and administered medications.

Our unit of analysis is the anesthesia case. without patient identifiers, repeated anesthestics provided to the same individual could not be identified and therefore such cases were analyzed independently. In Quarter 4 of 2014, NACOR contained about one million anesthesia cases with complete electronic anesthesia records. Our AQI created customized dataset  contained 441645 cases (superset) where intra-operative antimetics utlization was electronically accessible; antiemetics were utilized in 234453 cases.

Besides patient demographics and American Association of Anesthesiology risk classification, provider indentifier, institution and location, procedure codes and other case characteristics contained in  NACOR's PUF, our customized dataset contained for each anesthesia case additional information indicating 

1. which antimetics if any was administered [dexamethasone, droperidol, ondansetron and/or phenergan] {superset with 441645 cases}

2. Median income based on patient’s zip rounded to 1000, {subset with 218580 cases contained in the superset},

3. Generic and detailed insurance information, {subset with 176334 casescontained in the superset}.

We described the characteristic of the superset and subsets of the NACOR dataset forming the bases of our analysis, i.e. anesthesia records with complete information on the administration of antiemetic prophylaxis. We explored the bivariate associations between the dichotomous  outcome variable antiemetic utlization and the independent variables describing patients, procedures & providers; patient characteristics included medical insurance status (our primary predictor of interest), patient age, gender, American Association of Anesthesiology risk classification. Neither race nor ethnicity was recorded in NACOR. We reported procedure types and indications (Billing code, modifiers, indication ICD code). Provider characteristics include information on the anesthetist [nurse anesthetist versus resident versus attending alone] and institutional data [geographic location, academic versus private versus government institution]. 

We used parametric tests where the assumptions of normality did not seem violated and non-parametric test, where graphical or statistical tests suggested possible violations of the underlying assumptions. We reported proportions, mean and standard deviation or the median and the interquartile range (as appropriate for the distribution of values observed for each parameter) and indicated the statistical test used in the table of characteristic of patients. 

## Statistical analysis

We fit a Bayesian and a classical  logistic regression in the subset of anesthesia cases in the National Anesthesia Clinical Outcomes Registry (NACOR) with information on medical insurance status and intra-operative antiemetic administration. Administration of antiemetic prophylaxis is the primary outcome and medical insurance status is the ordinal predictor, controlling for patient & provider characteristics. We will used three different approaches to analyse the data:

1. a logistic regression model, 
2. a multivariate stratification and
3. a hierarchical Bayesian model.  

## Software used

we used the statistical software Stata\textsuperscript{\textregistered} for the regression analysis[@Stata_2014]. We used R, the public domain statistical software package[@R_software] and the probabilist programming Software Stan[@Stan_Software_2014] to implement the hierarchical Bayesian model. R and Stan are available under the General Public License of the Free Software Foundation[@GNU_1991] at no cost. We used the software package shinyStan[@shinystan-software_2015] for graphical exploration of model convergence, to generate the contrasts to compare commerical versus medicaid and for posterior predicitive checking.

## Modeling

We investigated the effect of medical insurance as predictor on the administration of antiemetic medication as primary outcome, controlling for potential confounders like patient characteristics, provider characteristics and procedure type and indication. The predictor insurance status is an ordinal variable; possible values are ordered from highest insurance coverage, i.e. private insurance, Health Maintenance Organization, medicare, medicaid, to the lowest no medical insurance reported. Our outcome is dichotomous, antiemetic prophylaxis administered or not. Our unit of analysis is the anesthesia case, not the patient. We focused our analysis on the most frequent procedures performed. We considered findings statistically significant if the p-value was less than the alpha of 0.05.Patients may have several operations, each generating one anesthesia record; patients may hence be counted several times. We will only consider antiemetic prophylaxis administered during anesthesia care, either in the operating room or while dropping off the patient in the recovery unit, but before sign-out to the recovery staff. 

### Classical logistic regression
We fit a classical logistic regression model. Besides insurance status as our primary outcome, we decided a priori to include gender and age in model, because both have been previously shown to be risk factors for PONY and are hence considered indications for antiemetic prophylaxis; as such they may act as confounders of associations. In addition, we  chose those independent variables for the initial model that show a statistically significant association in the bivariate analysis. We used stepwise backward elimination starting from our initial model based on the likelihood ratio test with a cutoff at 0.05 to eliminate independent variables from the model. For each model eliminated we confirmed that the given variable was not a confounder for the present model. We used a change in the beta coefficient of larger than 20% as our cutoff to determine if a variable is considered a confounder. We determined the correct functional form and explore potential violations of the assumptions of linearity. We ran locally weighted regression of yvar on xvar (and examined the graph for all independent variables in our final model), for a graphical assessment of potential violations of the assumption of linearity. We tested for the correct functional form, fitting fractional polynomials as part of our final logistic regression model. We  examined if the addition of a polynomial improves the model significantly. We explored the potential interaction between the independent variables age and gender in a simple logistic regression model; a cut of for our likelihood ratios test was at an alpha level of 0.05. We examined the goodness of fit with the Hosmer-Lemeshow goodness of fit test. We performed a sensitivity analysis of our model assumptions and choices.

### Stratified analysis
We calculated odds ratios for the association between insurance status and antiemetic administarion with the data stratified by gender, age and other demographics and case charactersitics.
 

### Bayesian hierarchical model
We build a hierarchical Bayesian model. We present the [Bayesian Model Code] in the appendix. We modeled the outcome as a dichtomous event $y_a$ for each anesthetic given, our unit of analysis

$$
  \begin{aligned} 
  y_{a} &\sim& Bernoulli(\mu_{a})\\
  \mu_a &\sim& inverse\_logit(\gamma[g_a] + x_a*beta)))
   \end{aligned}
$$

where $\mu_a$ is the probability of an event (ondansetron administration) in the a^th^ anesthetic, $g_a$ indicates the institution (where the anesthetic was given to the patient), $\gamma$ is the random effect each institution has on the administration of the antiemetic in a given case, $\beta$ the the coefficient of effect for the contrast between medicaid and commerical insurance and x_a the indicator for the insurance status.

Cases are nested in providers and these will be nested in institutions. (There are no patient identifiers to track repeated operations to the same patient.) We used uninformative priors for the main effect of insurance status on antiemetic prophylaxis. We performed a sensitivity analysis investigating the appropriateness of our model assumptions. 

We compared the results of our three models (logistic regression, stratified analysis and Bayesian hierarchical model) to confirm the robustness of our findings regardless of the choice of model or the statistical approach chosen.

-----

\newpage

# Results

## Description of the dataset

Our analysis is focused on the subset of the above subpopulation with information on both antiemetic usage *and* insurance status. The prevalence of ondansetron administration for different insurance status is  described in [Table 2 Insurance status versus antiemetic prophylaxis]. Clearly,  Patients who have commerical insurance are more likely to receive antiemetic prophylaxis than those who have medicaid.

Unfortunately, only 441645 cases in the Public User File (PUF) contained detailed information on medications administered during anesthesia administration at the end of 4th quarter 2014. Six unique institutions reported insurance status and antiemetic utilization for 176,334 anesthesia cases, out of the 12 million cases in NACOR 2014. Limiting this set further to cases with information on medical insurance, the dataset of our study contained 176334 anesthesia cases.The reporting institutions were mostly university hospitals or medium to large community hospitals, mostly in the Southern United States . Our dataset contained no cases from the Midwest or the West of the Unites States. Anesthesia was provided between 2010 and 2014. 

### Population
The demograpics of the subpopulation in the NACOR database with information on antiemetic administration are described in [Table 1 Population characteristics]. 42 percent of anesthetics were administered to male patients. Patients’ age ranged from newborn to 90 years of age with a median of 54 (Interquartile range (IQR) between 35 and 67 years). Most patients were classified as ASA class 2, 3 or 4 (39, 37 and 16 %, respectively) with only 7 percent in class 1 or 5.  62 percent were outpatients among the 64% of cases where this information was available. For 25865 cases the insurances was reported as medicaid, for another 97443 as medicare for the remaining 51441cases mostly as commercial insurance with 1585 self-insured cases. Some antiemetic was administered in 55 percent of all cases. 

### Bivariate comparison of demographic charactersitics
We found  statistically significant differences (p<0.001) in all demographic bivariate comparisons; they suggests that patients who received ondansetron (Zofran\textsuperscript{\textregistered}) were on average younger and more likely to be female, to have a slighly longer procedure, but as an outpatient, less likely to have undergone emergency surgery, more likely to live in the South in a zip code with higher mean income and smaller population size. 

## Association between insurance status and antiemetic administration

### Logistic regression analysis
We show the results of our final logistic regression model in the [Table 3 Logistic regression]. Being on medicaid or medicare, compared to having commerical insurance, drastically reduces the odds of receiving ondansetron during anesthesia. For the average patient on medicaid or medicare, the odds of receiving ondansetron for anti-emetic prophylaxis are 0.6 compared to a patient with commerical insurance with otherwise similar characteristics, even after controlling for age and gender, case duration, median income and population in the patient's home zip code. 

### Stratified analysis

We performed a stratified analysis, calculating the odds ratios for receiving ondansetron for patients on medicaid insurance versus commercial insurance. We show the results in the [Table 4 startified odds ratios] for the stratification by gender, older age, general anesthesia, outpatient surgery or the presence of a board certified anesthesiologist.  Stratification did change the odds ratios. However in all strata,  medicaid insurance status was associated with reduced odds of receiving antiemetic prophylaxis suggesting that the strong assoication between insurance status and antiemetic prophylaxis holds accross the board.

### Hierachical Bayesian model

[Table 5  Bayesian model] shows the parameter estimates for our simple Bayesian hierarchical model in a table. This Bayesian model only controlled for the effect of the institution as a random effect. Also in this model, having medicaid versus commerical insurance  was associated with receiving anti-emetic prophylaxis with an odds ratio of 0.6 with a narrow credible 95% interval [0.59, 0.62] . Remarkably, the random effects estimated for the different institutions in which the surgery took place vary considerably, suggesting that healthcare disparities may be endemic, or locally more or less pronounced.

#### Our Bayesian model converged. 
The convergence of our Bayesian model was confirmed by looking at traceplots and the Gelman Rubin statistics[@R_hat_Gelman_1992], shown for selected parameters of our Bayesian model.

## Sensitivity Analysis

The strong and statistically significant association (p < 0.001) between insurance status and the odds of receiving antiemetic prophylaxis remained unchanged after stratification to control for patient characteristics like gender and age. Our inferences were invariant to the statistical approach (Bayesian versus classical frequentist analysis) we used and bore out in the subset with data for multivariate logistic regression as well as in stratified analysis.  

# Discussion (1,500 word limit)

## Summary of the findings

Bivariate analysis, multivariate logistic regression and our Bayesian hierarchical model all showed a strong and statistically significant association between insurance status and the odds of receiving antiemetic prophylaxis; for medicaid versus commercially insured patients, the odds ratio of receiving the antiemetic ondansetron is 0.63 in our final logistic regression model with a 95% confidence interval of [0.62, 0.66] with similar Bayesian inferences. 

## Significance
To single out the effect of anesthesia providers in the perioperative process, can be perplexing[@Cook_22353147; @Silber_23719571]; we conceptualize the many potential confounders as (1) patient preoperative risk factors, (2) procedure related confounders, (3) institutional or geographic factors [@Cooper_22566148] (4) systematic and institutional bias, (5) provider behavior. 

## Strength
The size of the subset of NACOR we studied likely makes this the largest study of healthcare disparities in anesthesia undertaken to date. This increased our power to detect an association between insurance status and antiemetic prophylaxis. Controlling for likely confounders decreases the chance that the association is spurious.

## Impact
Demonstrating health care disparities in such a large dataset in anesthesia is novel. The fact that the surrogate marker of anesthesia quality is exclusively in the domain of the anesthesia will likely make a greater impression on the anesthesia community than an intervention or marker for which anesthesiologists are only partially involved or responsible for, like procedure time [@Silber_23719571]. Our work clearly puts the onus on the individual provider as opposed to system bias due to regulations or geographic variations. Our findings may hopefully lead to a somber assessment of healthcare disparities in anesthesia by all stakeholders, and appropriate counter measures. The anesthesia community has risen to the opportunity in the past and provided leadership in provision of equitable and safe care to all our patients[@Gaba_10720368].

## Limitations

### External and internal validity of our findings 

#### Our population may not be representative
Our sample is likely not representative of the general patient population undergoing anesthesia in the US, but oversampled academic institutions and the North East; we only have data from the institutions that uploaded full electronic anesthesia records; these institutions and their anesthesia practices are likely different from the average practice, e.g. academic practice is very different from private practice, with the later presumably catering more to outcomes patients perceive as important, PONY being one of them. 

#### Selection bias may limit our conclusions. 
Not all providers and institution use electronic anesthesia records and if they do they may not upload the data to NACOR, e.g. for regulatory or privacy reasons. Will certain providers with greater likelihood to discriminate on insurance status be less likely to have electronic medical records or to upload their datasets? This possibility may limit the generalizability of our findings as this may result in selection bias. Hence, our analysis should be repeated with a broader sample  as soon as additional patients data are uploaded to NACOR.

#### Information bias may result in under- or overestimation of the observed effect. 
Providers may have forgotten to record the administration of antiemetic medication, but still have administered the prophylaxis. They may have recorded the dose, but might still have failed to administer the prophylaxis. Both instances of misclassification of outcome (antiemetic prophylaxis administration) could lead to an over- or underestimation of healthcare disparities as we defined them. This distortion of the effect would be larger if the misclassification were not completely at random but differential depending on our predictor of interest (health insurance status). Because antiemetic prophylaxis is generally not considered critical or relevant for billing and malpractice, malicious intent to falsify the record is unlikely to lead to numerous differential misclassifications. Sloppy failure to record the medication, while actually giving it, seems a more likely occurrence. However, poor record keeping is likely associated with poor anesthesia quality and hence this misclassification would not invalidate the conclusions we draw from our analysis.

#### Our conclusions hinge on our assumption. 
If antiemetic prophylaxis is not a good indicator of anesthesia quality, then we cannot conclude that healthcare disparities exist in anesthesia, even if we find that insurance status predicts antiemetic administration. Likewise, if or as soon as antiemetic prophylaxis is used widely as such a marker, bias will be introduced in the documentation and administration of these agents and they may no longer genuinely reflect quality. Analogously, insurance status while closely linked to race and socioeconomic status is only one predictor of healthcare disparities and we do not know what might cause providers to differentiate based on the former versus the latter and/or if the same causal pathways are involved.

### Sensitivity analysis   

#### Each of our statistical approaches has its limitations. 
Stratified analysis only controls for the specific variable by which the stratification is done and doing substratification quickly becomes overwhelming. Logistic regression does not control well for confounding when the probability of effect is around 0.5. Hierarchical models lead to shrinkage, which makes for a better fit of the model to the data, but this may lead, among other issues, to overfitting.

#### Our results were invariant to the statistical approach.
If we used stratification, hierarchical Bayesian modeling or classical frequentist logistic regression, we came to the same inferences. But while we controlled for some important patient characteristics, we still need to consider surgery and provider as confounders and implement a full Bayesian hierarchical model. Certain surgical interventions are (believed) to be associated with a higher incidence of PONY and in some,is very undesiable, (e.g. because  the wretching concomittant with PONY can put undue stress on the surgical closure.) This can lead to confounding, if providers selectively administer antiemetics in these cases and surgical intervention and insurance status are associated. 


## Continous improvement is the ideal in healthcare
In the process of qulity improvement, we typically go through the four stages described by Don Berwick[@Berwick_2909878]:

1. Stage I: “The data are wrong….”
2. Stage II: “The data are right, but it's not a problem…”
3. Stage III: “The data are right, it's a problem, but it's not my problem…”
4. Stage IV: “The data are right, it's a problem, it's my problem…”

We hope to convince anesthesiologists with the presented strong and consistent association between insurance status and antiemetic prophylaxis regardless of the statistical approach, that the findings are solid. We think there is a clear argument that the described assoicatation suggest healthcare disparity for which anesthesia providers are accountable. The next step is to work to convince the community that this is our problem and that we can and should do something about this. Continous improvement is the ideal in healthcare[@Berwick_2909878] and we as anesthesiologist have a tradition to defend to be the leaders in continued improvement addressing individual performance as well as systems to achieve the best of care for all of our patients[@Gaba_10720368]

# Conclusion: 

Our analysis of the association between insurance status (as a marker of patient socioecomonic status) and antiemetic administration (as a marker of anesthesia quality provided by the individual anesthesiologist) point to possible unappreciated healthcare disparities in anesthesia at the provider level which are worrisome and deserve further investigation and vigorous countermeasures.


\newpage

# Tables

## Table 1 Population characteristics {#Table1}

### first Table 1

```{r make_first_table1, eval=FALSE}
Table <- read.csv("C:/Users/Micheal/Dropbox/Professional/AQI/Yeti/AQI Yeti/Table.csv")
save(Table1, file="Tables/Table")
```

```{r,eval=FALSE}
# C:\Users\Michael Andreae\Dropbox\Professional\AQI\Robert\tables for abstract
library(knitr)
load("Tables/Table")
kable(Table[1:9,1:4], row.names=FALSE)
```


### second Table 1

```{r second_table1}
require(knitr)
load("Tables/Table1")
kable(Table1)
```
Table 1 Population characteristics describes the subpopulation in the NACOR database with information on antiemetic administration. We found  statistically significant differences (p<0.001) in all comparisons (age, gender, income in the patients zip code and case characteristics like outpatient procedure, emergency procedure and case duration.



## Table 2 Insurance status versus antiemetic prophylaxis{#Table2}

```{r,eval=FALSE}
kable(Table[10:14,1:4], row.names=FALSE)
```

Table 2 describes the subset of patients (among the subpopulation from Table 1 Population characteristics with information on antiemetic administation) who have data on insurance status. Clearly  Patients who have commerical insurance are more likely to receive antiemetic prophylaxis than those who have medicaid.

## Table 3 Logistic regression {#Logistic_table}

```{r,eval=FALSE}
# AQI_stata_data_analysis.Rmd
rm(list=ls())
require(knitr)
load("table_my_logistic2.Rdata")
kable(table_my_logistic2, digit=3, row.names=TRUE)
```

The results of our logistic regression, presented in a table as odds ratios with the coresponding 95% confidence intervals, show that being on medicaid or medicare, compared to having commerical insurance, drastically reduces the odds of receiving ondansetron during anesthesia.

## Table 4 startified odds ratios 

```{r}
rm(list=ls())
require(knitr)
load("Tables/LOR_table.Rdata")
kable(OR_table, digits=3)
```

This table shows the odds ratios for receiving ondansetron for patients on medicaid versus patients on commercial insurance after we stratified for various possible confounders. Stratifying for gender, age, general anesthesia, outpatient surgery or the presence of a board certified anesthesiologist did change the odds ratio. However in all strata,  medicaid insurance status was associated with reduced odds of receiving antiemetic prophylaxis suggesting that the association holds accross the board.


##Table 5  Bayesian model {#Bayesian_table}

```{r}
rm(list=ls())
require(knitr)
load("Tables/Bayesian_table.Rdata")
kable(Bayesian_table[,c(2,1,3,4,5)], digits=3 )
```

Having medicaid versus commerical insurance reduces the odds of receiving anti-emetic prophylaxis significantly in our Bayesian hierarchical model. This is clear from the estimated odds ratio of receiving ondansetron for the contrast of $\beta_1$ (medicaid) versus $\beta_3$ (commercial) shown with the corresponding 95% Bayesian credible intervals. Remarkably, the random effects estimated for the different institutions in which the surgery took place vary considerably, suggesting that healthcare disparities may be endemic, or locally more or less pronounced.

\newpage

# Appendix


\setlength{\parindent}{0ex}

\newpage

# Figure Legends

\newpage

# Figures

\newpage

# Supplemental Digital Content

## Bayesian Model Code

\newpage

# References