---
title: "AQI_ManuscriptTables"
author: "Michael Andreae"
date: "December 3, 2015"
output: pdf_document
---

```{r, global_options, echo=FALSE}
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

# Prepare

## Load Data to produce table

Load the complete case dataset produced with AQI_Import  on 3Dec2015 once from the source data file and save it in */Data* and clean workspace.

```{r, importData, eval=FALSE}
load("C:/Users/Micheal/Dropbox/Professional/AQI/AQI_Import/myAQI_cpt.Rdata")
save(ccAQI_cpt, myAQI_cpt, 
     file = "Data/ManuscriptData3Dec2015.Rdata")
rm(list=ls())
```

```{r, Libraries}
library(knitr)
library(dplyr)
```

```{r, loadData}
load("Data/ManuscriptData3Dec2015.Rdata")
str(ccAQI_cpt)
```

# Build Tables

## Table 1 Population characteristics

Table 1 Population characteristics describes the subpopulation in the NACOR database with information on antiemetic administration. We found statistically significant differences (p<0.001) in all comparisons (age, gender, income in the patients zip code and case characteristics like outpatient procedure, emergency procedure and case duration.

```{r, Table1}
attach(myAQI_cpt)

Variables <- c("n",               # all
               levels(sex), # gender
               "median [IQR)",
               levels(age_group), 
               levels(pay),
                "income",levels(incomeQ))


Total <- c(dim(myAQI_cpt)[1], 
           table(sex),
           median(age, na.rm=TRUE),
           table(age_group),
           table(pay),
           seq(1, length(Variables)-14))

Groups <- c(
  c("Population", "Gender", ""),
  c("Age", rep("",6)),
  c("Insurance", rep("",3)),
  c("Income", rep("",4))
            )

Table1 <- data.frame(Groups=Groups, Variables=Variables, Total=Total)

kable(Table1)

```