---
title: "AQI_ManuscriptTables"
author: "Michael Andreae"
date: "December 3, 2015"
output: pdf_document
---

```{r, global_options, echo=FALSE}
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

# Prepare

## Load Data to produce table

Load the complete case dataset produced with AQI_Import  on 3Dec2015 once from the source data file and save it in */Data* and clean workspace.

```{r, importData, eval=FALSE}
load("C:/Users/Micheal/Dropbox/Professional/AQI/AQI_Import/myAQI_cpt.Rdata")
save(ccAQI_cpt, myAQI_cpt, 
     file = "Data/ManuscriptData3Dec2015.Rdata")
rm(list=ls())
```

```{r, Libraries}
library(knitr)
library(dplyr)
```

```{r, loadData}
load("Data/ManuscriptData3Dec2015.Rdata")
str(ccAQI_cpt)
```

# Build Tables

\newpage

## Table 1 Population characteristics


```{r, Table1}
attach(myAQI_cpt)

Variables <- c("n",               # all
               levels(sex), # gender
               "median [IQR]",
               levels(age_group), 
               levels(pay),
                "median [IQR]",levels(incomeQ))


Total <- as.integer(c(dim(myAQI_cpt)[1], 
           table(sex),
           median(age, na.rm=TRUE),
           table(age_group),
           table(pay),
           median(income, na.rm = TRUE)*100000,
           table(incomeQ)
           ))

Groups <- c(
  c("Population", "Gender", ""),
  c("Age", rep("",6)),
  c("Insurance", rep("",3)),
  c("Income ($)", "-quartiles", rep("",3))
            )

Table1 <- data.frame(Groups=Groups, Variables=Variables, Total=Total)

kable(Table1)

```

Table 1 Population characteristics describes the subpopulation in the NACOR database with information on antiemetic administration. We found statistically significant differences (p<0.001) in all comparisons (age, gender, income in the patients zip code and case characteristics like outpatient procedure, emergency procedure and case duration.

### Table by hand

|     |      |   |
|-----|------|---|
| all | 3442 |   |
| age | 4556 |   |

|     |      |   |
|-----|------|---|
|Population | n | 441645 |
|------------------|
|Gender | female | 249909 |
| | male | 191654 |
||||
| Age | median | `r median(age, na.rm=TRUE)` |


```{r}

medInc <- median(myAQI_cpt$age, na.rm=TRUE)
```


\newpage 

  Right     Left     Center     Default
-------     ------ ----------   -------
`r median(myAQI_cpt$age, na.rm=TRUE)`     12        12            `r median(myAQI_cpt$age, na.rm=TRUE)`
    123     123       123          123
      1     1          1             1

Table:  Demonstration of simple table syntax.

  Right     Left     Center     Default
-------     ------ ----------   -------
     12     12        12            12
    123     123       123          123
      1     1          1             1

Table:  Demonstration of simple table syntax.