---
output:
  word_document:
    reference_docx: mytables.docx
---


```{r, global_options, echo=FALSE}
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

***

## Table 2. Predictors of antiemetic administration {#Table2}

---

```{r make_first_table1, eval=FALSE}
Table <- read.csv("C:/Users/Micheal/Dropbox/Professional/AQI/Yeti/AQI Yeti/Table.csv")
names(Table) <- c("Variable", "Total", "No Ondansetron", 
               "Ondansetron", "P.value")
save(Table, file="Table.Rdata")
```

```{r,eval=TRUE}
# C:\Users\Michael Andreae\Dropbox\Professional\AQI\Robert\tables for abstract
library(knitr)
load("Table.Rdata")
kable(Table[1:9,1:4], row.names=FALSE)
```

Table 2 describes the demographic characteristics predicting ondansetron use in the NACOR superset with electronic data on antiemetic administration. There seems to be a slight preponderance of ondansetron use in female patients, (who are at higher risk for postoperative nausea and vomiting) and in patients living in a zip code with a higher median income. 