---
output:
  word_document:
    reference_docx: mytables.docx
---


```{r, global_options, echo=FALSE}
suppressWarnings(library(knitr))
# set options
opts_chunk$set(fig.width=6, fig.height=4, fig.path='Figs/',
               echo=FALSE, warning=FALSE, message=FALSE)
eval_everything=FALSE
```

***

## Table 3. Insurance Status Versus Antiemetic Prophylaxis {#Table3}

---

```{r,eval=TRUE}
library(knitr)
load("Table.Rdata")
kable(Table[10:14,1:4], row.names=FALSE)
```

Table 3 describes the association of insurance status with antiemetic adminstration in the NACOR case subset with data on 176334 anesthetic cases antiemetic adminstration *and* insurance status. Clearly  patients  with commercial insurance are more likely to receive antiemetic prophylaxis than those who are insured through medicaid or medicare. (This association was statistically highly significant in a Chi-square analysis with a p value < 0.001, in brackets the insurances in percent of the cases with and without ondansetron use.)